Towards Practical First-Order Model Counting

### 2024-08-11 doc/paper/paper.tex
Made it fit

\newcommand{\crefrangeconjunction}{--}
\crefname{line}{line}{lines}
\crefname{fact}{Observation}{Observations}
\crefname{assumption}{Assumption}{Assumptions}
\crefalias{enumi}{type}
\crefname{type}{Type}{Types}
\creflabelformat{type}{#2\textup{#1}#3}
\crefalias{enumi}{step}
\crefname{step}{Step}{Steps}
\creflabelformat{step}{#2\textup{#1}#3}
\crefalias{clause}{equation}
\crefname{clause}{Clause}{Clauses}
\creflabelformat{clause}{#2\textup{(#1)}#3}
\crefalias{formula}{equation}
\crefname{formula}{Formula}{Formulas}
\creflabelformat{formula}{#2\textup{(#1)}#3}

@@

\emph{First-order model counting} (FOMC) is the task of counting the number of
models of a sentence in first-order logic over some given domain(s). The
(symmetric) weighted variant of this problem, known as WFOMC, seeks to compute
the total weight of the models~\cite{DBLP:conf/ijcai/BroeckTMDR11}. WFOMC
relates to its propositional predecessor weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08} and, more conceptually, to other
attempts to unify logic and probability such as probabilistic
logic~\cite{DBLP:journals/ai/Nilsson86}, fuzzy
logic~\cite{novak2012mathematical}, and graded logics~\cite{vsaletic2024graded}.
Moreover, WFOMC is a key approach to \emph{lifted (probabilistic) inference},
which aims to compute probabilities more efficiently by leveraging symmetries
inherent in the problem description~\cite{DBLP:conf/ecai/Kersting12}.

@@

Lifted inference is an active area of research, with recent work in domains such
as constraint satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23} and
probabilistic answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}.
WFOMC has been used for inference on probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. By considering domains of
increasing sizes, the FOMC of a formula can be seen as an integer sequence. FOMC
algorithms have been utilised for discovering new
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23} as well as
conjecturing~\cite{DBLP:conf/ilp/BarvinekB0ZK21} and
constructing~\cite{DBLP:conf/kr/DilkasB23} recurrence relations and other
recursive structures that describe these sequences. Additionally, FOMC
algorithms have been extended to perform
\emph{sampling}~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}.

@@

The complexity of FOMC is typically characterised in terms of \emph{data
  complexity}. This involves fixing the formula and determining whether an
algorithm exists that can compute the FOMC in time polynomial with respect to
the domain size(s). If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}.
\citeauthor{DBLP:conf/pods/BeameBGS15}~\shortcite{DBLP:conf/pods/BeameBGS15}
demonstrated the existence of an unliftable formula with three variables. It is
also known that all formulas with up to two variables are
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}. The liftable
fragment of formulas with two variables has been expanded with various
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23}, counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21} and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

@@

There are various FOMC algorithms with different underlying principles. Perhaps
the most prominent class of FOMC algorithms is based on \emph{first-order
  knowledge compilation} (FOKC). In this approach, the formula is compiled into
a representation (such as a circuit or graph) by iteratively applying
\emph{compilation rules}. This representation can then be used to compute the
FOMC for any combination of domain sizes. Algorithms in this class include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent extension
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles formulas into
circuits, while the latter compiles them first to graphs and then to (algebraic)
equations. Another FOMC algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, is based on cell enumeration.
Other algorithms utilise local search~\cite{DBLP:journals/pvldb/NiuRDS11},
junction trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.

@@

The recently proposed \textsc{Crane} algorithm marked significant progress in
handling formulas beyond the capabilities of \textsc{FastWFOMC} and
\textsc{ForcLift}, yet it fell short in critical aspects. \textsc{Crane} was
incomplete in that it could only construct function definitions, leaving users
to manually evaluate these functions to obtain model counts. This limitation
prevented \textsc{Crane} from serving as a convenient black box solution for
FOMC. Furthermore, it introduced recursive functions without defining necessary
base cases, adding further complexity to the user.

In this work, we present \Cranetwo{}, the breakthrough that not only addresses
these gaps but also pushes the boundaries of scalability to unprecedented
levels. With \Cranetwo{}, we eliminate the manual steps, providing a fully
automated FOMC algorithm. Moreover, \Cranetwo{} can scale to domain sizes over
\num{500000} times larger than the previous state of the art, redefining what is
possible in the field of first-order model counting.

@@

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for the formulas
from \cref{step2} above required to preserve the correct model count. Next,
\cref{sec:cpp} elucidates how function definitions encoding a solution to a FOMC
problem are compiled into C++ programs. Note that a solution to FOMC that uses
compilation to C++ has been considered before~\cite{DBLP:conf/kr/KazemiP16},
however, the extent of formulas that could be handled was limited. Finally,
\cref{sec:experiments} presents experiments results comparing \Cranetwo{} with
other state-of-the-art FOMC algorithms.

@@

In \cref{sec:logic}, we provide a summary of the basic principles of first-order
logic. Then, in \cref{sec:threelogics}, we formally define (W)FOMC and discuss
the distinctions between three variations of first-order logic that are utilised
for FOMC.\@ Finally, in \cref{sec:algebra}, we introduce the terminology used
to describe the output of the original \textsc{Crane} algorithm, i.e., functions
and equations that define them.

@@

In this section, we review the basic concepts of first-order logic as they are
used in FOKC algorithms. There are two key differences between the logic used by
such algorithms and the logic supported as input. First,
Skolemization~\cite{DBLP:conf/kr/BroeckMD14} is employed to eliminate
existential quantifiers by introducing additional predicates. Note that
Skolemization here is different from the standard Skolemization procedure that
introduces function symbols~\cite{DBLP:books/daglib/0030198}. Second, the input
formula is rewritten as a conjunction of clauses, each of which is in
\emph{prenex normal form}~\cite{hinman2018fundamentals}.

A \emph{term} is either a variable or a constant. An \emph{atom} is either
\begin{enumerate*}[label=(\roman*)]
  \item $P(t_{1}, \dots, t_{m})$ for some predicate $P$ and terms
  $t_{1}, \dots, t_{m}$ (written as $P(\mathbf{t})$ for short) or
  \item $x=y$ for some terms $x$ and $y$.
\end{enumerate*}
The \emph{arity} of a predicate is the number of arguments it takes, i.e., $m$
in the case of predicate $P$ above. We write $P/m$ to denote a predicate
together with its arity. A \emph{literal} is either an atom (i.e., a
\emph{positive} literal) or its negation (i.e., a \emph{negative} literal). An
atom is \emph{ground} if it contains no variables, i.e., only constants. A
\emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\text{.}\forall x_{2} \in \Delta_{2}\dots\text{}\forall x_{n} \in \Delta_{n}\text{.}\phi(x_{1}, x_{2}, \dots, x_{n})$,
where $\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). We say that a clause is a
\emph{(positive) unit clause} if
\begin{enumerate*}[label=(\roman*)]
  \item there is only one literal with a predicate, and
  \item it is a positive literal.
\end{enumerate*}
Finally, a \emph{formula} is a conjunction of clauses. Throughout the paper, we
use set-theoretic notation, interpreting a formula as a set of clauses and a
clause as a set of literals.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in FOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

@@

Compared to \FO{}, \Ctwo{} and \UFO{} lack support for:
\begin{enumerate*}[label=(\roman*)]
  \item constants,
  \item the equality predicate,
  \item multiple domains, and
  \item formulas with more than two variables.
\end{enumerate*}
The advantage that \Ctwo{} brings over \FO{} is the inclusion of \emph{counting
  quantifiers}. That is, alongside $\forall$ and $\exists$, \Ctwo{} supports
$\exists^{=k}$, $\exists^{\le k}$, and $\exists^{\ge k}$ for any positive
integer $k$. For example, $\exists^{=1} x\text{. }\phi(x)$ means that there
exists \emph{exactly one} $x$ such that $\phi(x)$, and $\exists^{\le 2} x\text{.
}\phi(x)$ means that there exist \emph{at most two} such $x$. \UFO{}, on the
other hand, does not support any existential quantifiers but instead
incorporates \emph{(equality) cardinality constraints}. For example, $|P| = 3$
constrains all models to have \emph{precisely three positive literals with the
  predicate $P$}.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains (which
  may not be distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets, satisfying the following conditions:
  \begin{enumerate*}[label=(\roman*)]
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate*}
  (In practice, we typically only specify the size of each domain.) A
  \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of ground
  literals defined by adding to $M$ either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$.
\end{definition}

\begin{remark}
  The distinctness of domains is important in two ways. First, in terms of
  expressiveness, clauses such as $\forall x \in \Delta\text{. }P(x, x)$ are
  valid if predicate $P$ is defined over two copies of the same domain and
  invalid otherwise. Second, having more distinct domains makes the problem more
  decomposable for the FOKC algorithm. With more distinct domains, the algorithm
  can make assumptions or deductions about, e.g., the first domain of predicate
  $P$ without worrying how (or if) they apply to the second domain.
\end{remark}

While the focus of this work is on FOMC, we still define the weighted variant of
the problem as Skolemization relies on weights even for unweighted FOMC.

\begin{definition}[WFOMC Instance]\label{def:instance}
  A \emph{WFOMC instance} comprises of:
  \begin{enumerate*}[label=(\roman*)]
    \item a formula $\phi$ in \FO{},
    \item two (rational) \emph{weights} $w^{+}(P)$ and $w^{-}(P)$ assigned to
    each predicate $P$ in $\phi$, and
    \item $\sigma$ as described in \cref{def:model}.
  \end{enumerate*}
  Unless specified otherwise, we assume all weights to be equal to 1.
\end{definition}

\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Given a WFOMC instance $(\phi, w^{+}, w^{-}, \sigma)$ as in
  \cref{def:instance}, the \emph{(symmetric) weighted first-order model count}
  (WFOMC) of $\phi$ (with respect to $\sigma$, $w^{+}$, and $w^{-}$) is
  \begin{equation}\label{eq:wfomc}
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \end{equation}
  where the sum is over all models of $\phi$.
\end{definition}

@@

We write $\expr{}$ to represent an arbitrary algebraic expression. It is
important to note that some terms have different meanings in algebra compared to
logic. In algebra, a \emph{constant} refers to a non-negative integer. Likewise,
a \emph{variable} can either be a parameter of a function or a variable
introduced through summation, such as $i$ in the expression
$\sum_{i=1}^{n} \expr$. A (function) \emph{signature} is
$f(x_{1}, \dots, x_{n})$ (or $f(\mathbf{x})$ for short), where $f$ represents an
$n$-ary function, and each $x_{i}$ represents a variable. An \emph{equation} is
$f(\mathbf{x}) = \expr{}$, with $f(\mathbf{x})$ representing a signature.

@@

Before describing the main contribution of this work, let us review the key
aspects of FOKC as realised by \textsc{Crane}. The input formula is compiled
into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
$\mathcal{E}$ can contain any number of functions, one of which (denoted by $f$)
represents the solution to the FOMC problem. To compute the FOMC of the formula
for particular domain sizes, $f$ must be evaluated with those domain sizes as
arguments. $\mathcal{D}$ records this correspondence between function arguments
and domains.

@@

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we use
the FOKC algorithm of the original \textsc{Crane} to compile the formula into
the three components: $\mathcal{E}$, $\mathcal{F}$, and $\mathcal{D}$. After
some algebraic simplification (explained at the end of this section),
$\mathcal{E}$ is passed to the \FindBaseCases procedure (explained in
\cref{sec:identifying}), which identifies the base cases that need to be
defined. For each base case $f(\mathbf{x})$, we retrieve the logical formula
$\mathcal{F}(f)$ associated with the function name $f$ and simplify it using the
\Propagate procedure (explained in detail in \cref{sec:simplifying}). We do this
by iterating over all indices of $\mathbf{x}$, where $x_{i}$ is a constant, and
using \Propagate to simplify $\psi$ by assuming that domain $\mathcal{D}(f, i)$
has size $x_{i}$. Finally, on \cref{line:final}, \CompileWithBaseCases recurses
on these simplified formulas and adds the resulting base case equations to
$\mathcal{E}$. \Cref{example:overall} below provides more detail.

@@

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Compile. (We omit the
  definition of $\mathcal{F}$ as the formulas can get a bit verbose.) Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we obtain the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, resulting in $g(l, 0) = 1$.
\end{example}

>>>>

\newcommand{\crefrangeconjunction}{--}
\crefname{line}{line}{lines}
%\crefname{fact}{Observation}{Observations}
\crefname{assumption}{Assumption}{Assumptions}
\crefalias{enumi}{type}
\crefname{type}{Type}{Types}
\creflabelformat{type}{#2\textup{#1}#3}
\crefalias{enumi}{step}
\crefname{step}{Step}{Steps}
\creflabelformat{step}{#2\textup{#1}#3}
\crefalias{clause}{equation}
\crefname{clause}{Clause}{Clauses}
\creflabelformat{clause}{#2\textup{(#1)}#3}
\crefalias{formula}{equation}
\crefname{formula}{Formula}{Formulas}
\creflabelformat{formula}{#2\textup{(#1)}#3}

@@

\emph{First-order model counting} (FOMC) is the task of counting the number of
models of a sentence in first-order logic over some given domain(s). The
weighted variant of this problem, known as WFOMC, seeks to compute the total
weight of the models~\cite{DBLP:conf/ijcai/BroeckTMDR11}. WFOMC is related to
its propositional predecessor weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08} and other attempts to unify logic
and
probability~\cite{DBLP:journals/ai/Nilsson86,novak2012mathematical,vsaletic2024graded}.
It is also a key approach to \emph{lifted inference}, which aims to compute
probabilities more efficiently by leveraging symmetries in the
problem~\cite{DBLP:conf/ecai/Kersting12}.

@@

Lifted inference is an active area of research, with recent work in domains such
as constraint satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23} and
probabilistic answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}.
WFOMC has been used for inference on probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. FOMC algorithms have been
utilised for discovering new integer
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}, and for
conjecturing~\cite{DBLP:conf/ilp/BarvinekB0ZK21} and
constructing~\cite{DBLP:conf/kr/DilkasB23} recurrence relations and other
recursive structures that describe these sequences. FOMC algorithms have also
been extended to perform
\emph{sampling}~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}.

% By considering domains of increasing sizes, the FOMC of a formula can be seen
% as an integer sequence.

@@

The complexity of FOMC is typically characterised in terms of \emph{data
  complexity}. If there is an algorithm that can compute the FOMC of a formula
in polynomial time with respect to the domain size(s), that formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}.
\citeauthor{DBLP:conf/pods/BeameBGS15}~\shortcite{DBLP:conf/pods/BeameBGS15}
demonstrated the existence of an unliftable formula with three variables. It is
also known that formulas with up to two variables are
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}. The liftable
fragment of formulas with two variables has been expanded with various
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23}, counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21} and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

% This involves fixing the formula and determining whether an
% algorithm exists that can compute the FOMC in time polynomial with respect to
% the domain size(s).

@@

There are many FOMC algorithms with different underlying principles. Perhaps the
most prominent class of FOMC algorithms is based on \emph{first-order knowledge
  compilation} (FOKC). In this approach, the formula is compiled into a
representation (such as a circuit or graph) by applying \emph{compilation
  rules}. Algorithms in this class include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its extension
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. Another FOMC algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, is based on cell enumeration.
Other algorithms utilise local search~\cite{DBLP:journals/pvldb/NiuRDS11},
junction trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.

% This representation can then be used to compute the FOMC for any combination
% of domain sizes.

% The former compiles formulas into circuits, while the latter compiles them
% first to graphs and then to (algebraic) equations.

@@

The recently proposed \textsc{Crane} algorithm marked significant progress in
handling formulas beyond the capabilities of \textsc{FastWFOMC} and
\textsc{ForcLift}, yet it fell short in critical aspects. \textsc{Crane} was
incomplete since it could only construct function definitions, requiring users
to manually evaluate these functions to obtain model counts. This limitation
prevented \textsc{Crane} from serving as a convenient black box solution for
FOMC. Furthermore, it introduced recursive functions without defining necessary
base cases, adding further complexity to the users. In this work, we present
\Cranetwo{}, addressing these gaps and pushing scalability to unprecedented
levels. Unlike its predecessor, \Cranetwo{} is a fully automated FOMC algorithm,
capable of handling domain sizes over \num{500000} times larger than previous
algorithms.

@@

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds base cases for
recursive functions. \Cref{sec:smoothing} explains post-processing techniques to
preserve the correct model count. \Cref{sec:cpp} elucidates how function
definitions are compiled into C++ programs. Note that a solution to FOMC that
uses compilation to C++ has been considered
before~\cite{DBLP:conf/kr/KazemiP16}, however, the extent of formulas that could
be handled was limited. Finally, \cref{sec:experiments} presents experimental
results comparing \Cranetwo{} with other FOMC algorithms.

% by:
% \begin{enumerate*}[label=(\roman*)]
%   \item identifying a sufficient set of base cases for each function,
%   \item constructing formulas corresponding to these base cases,
%   and\label[step]{step2}
%   \item recursing on these new formulas.
% \end{enumerate*}

@@

In \cref{sec:logic}, we summarise the basic principles of first-order logic.
Then, in \cref{sec:threelogics}, we formally define (W)FOMC and discuss the
distinctions between three variations of first-order logic used for FOMC.\@
Finally, in \cref{sec:algebra}, we introduce the terminology used to describe
the output of the original \textsc{Crane} algorithm, i.e., functions and
equations that define them.

@@

In this section, we will review the basic concepts of first-order logic as they
are used in FOKC algorithms. There are two key differences between the logic
used by these algorithms and the logic supported as input. First,
Skolemization~\cite{DBLP:conf/kr/BroeckMD14} eliminates existential quantifiers
by introducing additional predicates. Please note that Skolemization here
differs from the standard Skolemization procedure that introduces function
symbols~\cite{DBLP:books/daglib/0030198}. Second, the input formula is rewritten
as a conjunction of clauses, each in \emph{prenex normal
  form}~\cite{hinman2018fundamentals}.

A \emph{term} can be either a variable or a constant. An \emph{atom} can be
either
\begin{enumerate*}[label=(\roman*)]
  \item $P(t_{1}, \dots, t_{m})$ for some predicate $P$ and terms
  $t_{1}, \dots, t_{m}$ (written as $P(\mathbf{t})$ for short) or
  \item $x=y$ for some terms $x$ and $y$.
\end{enumerate*}
The \emph{arity} of a predicate is the number of arguments it takes, i.e., $m$
in the case of the predicate $P$ mentioned above. We write $P/m$ to denote a
predicate along with its arity. A \emph{literal} can be either an atom (i.e., a
\emph{positive} literal) or its negation (i.e., a \emph{negative} literal). An
atom is \emph{ground} if it contains no variables, i.e., only constants. A
\emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
}\forall x_{2} \in \Delta_{2}\dots\text{ }\forall x_{n} \in \Delta_{n}\text{.
}\phi(x_{1}, x_{2}, \dots, x_{n})$, where $\phi$ is a disjunction of literals
that only contain variables $x_{1}, \dots, x_{n}$ (and any constants). We say
that a clause is a \emph{(positive) unit clause} if
\begin{enumerate*}[label=(\roman*)]
  \item there is only one literal with a predicate, and
  \item it is a positive literal.
\end{enumerate*}
Finally, a \emph{formula} is a conjunction of clauses. Throughout the paper, we
will use set-theoretic notation, interpreting a formula as a set of clauses and
a clause as a set of literals.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in FOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and an $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

@@

Compared to \FO{}, \Ctwo{} and \UFO{} lack support for
\begin{enumerate*}[label=(\roman*)]
  \item constants,
  \item the equality predicate,
  \item multiple domains, and
  \item formulas with more than two variables.
\end{enumerate*}
The advantage that \Ctwo{} brings over \FO{} is the inclusion of \emph{counting
  quantifiers}. That is, alongside $\forall$ and $\exists$, \Ctwo{} supports
$\exists^{=k}$, $\exists^{\le k}$, and $\exists^{\ge k}$ for any positive
integer $k$. For example, $\exists^{=1} x\text{. }\phi(x)$ means that there
exists \emph{exactly one} $x$ such that $\phi(x)$, and $\exists^{\le 2} x\text{.
}\phi(x)$ means that there exist \emph{at most two} such $x$. \UFO{}, on the
other hand, does not support any existential quantifiers but instead
incorporates \emph{(equality) cardinality constraints}. For example, $|P| = 3$
constrains all models to have \emph{precisely three positive literals with the
  predicate $P$}.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains. Let
  $\sigma$ be a map from the domains of $\phi$ to their interpretations as sets,
  satisfying the following conditions:
  \begin{enumerate*}[label=(\roman*)]
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate*}
  A \emph{structure} of $\phi$ is a set $M$ of ground literals defined by adding
  to $M$ either $P(\mathbf{t})$ or $\neg P(\mathbf{t})$ for every predicate
  $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$.
\end{definition}

% (In practice, we typically only specify the size of each domain.)

\begin{remark}
  The distinctness of domains is important in two ways. First, in terms of
  expressiveness, a clause such as $\forall x \in \Delta\text{. }P(x, x)$ is
  valid if predicate $P$ is defined over two copies of the same domain and
  invalid otherwise. Second, having more distinct domains makes the problem more
  decomposable for the FOKC algorithm. With distinct domains, the algorithm can
  make assumptions or deductions about, e.g., the first domain of predicate $P$
  without worrying how (or if) they apply to the second domain.
\end{remark}

While this work focuses on FOMC, we still define the weighted variant of the
problem as Skolemization relies on weights even for unweighted FOMC.

\begin{definition}[WFOMC instance]\label{def:instance}
  A \emph{WFOMC instance} comprises:
  \begin{enumerate*}[label=(\roman*)]
    \item a formula $\phi$ in \FO{},
    \item two (rational) \emph{weights} $w^{+}(P)$ and $w^{-}(P)$ assigned to
    each predicate $P$ in $\phi$, and
    \item $\sigma$ as described in \cref{def:model}.
  \end{enumerate*}
  Unless specified otherwise, we assume all weights to be equal to 1.
\end{definition}

\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Given a WFOMC instance $(\phi, w^{+}, w^{-}, \sigma)$ as in
  \cref{def:instance}, the \emph{(symmetric) weighted first-order model count}
  (WFOMC) of $\phi$ is
  \begin{equation}\label{eq:wfomc}
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \end{equation}
  where the sum is over all models of $\phi$.
\end{definition}

@@

We write $\expr{}$ to represent an arbitrary algebraic expression. It is
important to note that some terms have different meanings in algebra and
logic. In algebra, a \emph{constant} refers to a non-negative integer. Likewise,
a \emph{variable} can either be a parameter of a function or a variable
introduced through summation, such as $i$ in the expression
$\sum_{i=1}^{n} \expr$. A (function) \emph{signature} is
$f(x_{1}, \dots, x_{n})$ (or $f(\mathbf{x})$ for short), where $f$ represents an
$n$-ary function, and each $x_{i}$ represents a variable. An \emph{equation} is
$f(\mathbf{x}) = \expr{}$, with $f(\mathbf{x})$ representing a signature.

@@

Before describing the main contribution of this work, let us review the
essential aspects of FOKC as realised by \textsc{Crane}. The input formula is
compiled into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
$\mathcal{E}$ can contain any number of functions, one of which (denoted by $f$)
represents the solution to the FOMC problem. To compute the FOMC for particular
domain sizes, $f$ must be evaluated with those domain sizes as arguments.
$\mathcal{D}$ records this correspondence between function arguments and
domains.

@@

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we use
the FOKC algorithm of the original \textsc{Crane} to compile the formula into
the three components: $\mathcal{E}$, $\mathcal{F}$, and $\mathcal{D}$. After
some algebraic simplification, $\mathcal{E}$ is passed to the \FindBaseCases
procedure (see \cref{sec:identifying}). For each base case $f(\mathbf{x})$, we
retrieve the logical formula $\mathcal{F}(f)$ associated with the function name
$f$ and simplify it using the \Propagate procedure (explained in detail in
\cref{sec:simplifying}). We do this by iterating over all indices of
$\mathbf{x}$, where $x_{i}$ is a constant, and using \Propagate to simplify
$\psi$ by assuming that domain $\mathcal{D}(f, i)$ has size $x_{i}$. Finally, on
\cref{line:final}, \CompileWithBaseCases recurses on these simplified formulas
and adds the resulting base case equations to $\mathcal{E}$.
\Cref{example:overall} below provides more detail.

@@

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain. (We omit the definition of
  $\mathcal{F}$ as the formulas can get a bit verbose.) Then \FindBaseCases
  identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both cases,
  \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$ simplified by
  assuming that one of the domains is empty. In the first case, we recurse on
  the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$, where $S$ is a
  predicate introduced by Skolemization with weights $w^{+}(S) = 1$ and
  $w^{-}(S) = -1$. Hence, we obtain the base case $g(0, m) = 0^{m}$. In the case
  of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$} returns an empty formula,
  resulting in $g(l, 0) = 1$.
\end{example}
